---
title: "CEU-R-Tools-Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Link to my project on github

https://github.com/AttilaKrajko/CEU-R-Project

## CLEAR MEMORY

```{r}
rm(list = ls())
```


```{r}

library(nycflights13)
library(data.table)
library(dplyr)
library(ggplot2)
library(scales)
library(plyr)
library(lubridate)
library(tidyr)
library(class)
library(pander)
```
Read the data table. 

```{r}

dtflights <- data.table(flights)
dtairpoirts <- data.table(airports)
dtairlines <- data.table(airlines)
dtplanes <- data.table(planes)
dtweather <- data.table(weather)
str(flights)
?flights
```

##Flights departure delay
```{r}

plot(sort(flights$dep_delay), type="h", ylab="Departure Delay")
```

##Arrival Time delays in minute
```{r}
hist(dtflights$arr_delay, main = "Arrival Time Delays [in minutes]")
```


##Flights departure delay with histogram and density plot
```{r}
hist(flights$dep_delay, xlab="Departure Delay")

plot(density(flights$dep_delay, na.rm=TRUE))
```

##Flights departure delay with boxplot
```{r}
boxplot(log(flights$dep_delay - 
            min(flights$dep_delay,na.rm=TRUE)
            +1), ylab="Departure Delay")
```

##The average departure and arrival delays per destination
```{r}

dta <- dtflights[, .(departure = mean(dep_delay, na.rm = TRUE),
              arrival = mean(arr_delay, na.rm = TRUE)), by = dest]
ggplot(dta, aes(departure, arrival, label = dest)) + geom_text()
```

##Histogram of Flight Air time

```{r}
hist(dtflights$air_time, xlab = "Flight Air Time [in minutes]", main = "Histogram of Flight Air Time")
```


## Which destination has the lowest average delay from 'EWR'?
```{r}
dta <- dtflights[origin == 'EWR', .(delay = mean(arr_delay, na.rm = TRUE)), by = dest]
setorder(dta, delay)
head(dta)
dta[1]

```

##The average delay to all destinations from 'EWR
```{r}
ggplot(dta, aes(dest, delay)) + geom_bar(stat = 'identity') + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

##The number of flights per month
```{r}

ggplot(dtflights[, .N, by = month], aes(month, N)) + geom_bar(stat = 'identity')
```


##Heatmap on the number of flights per month and hour of the day
```{r}
ggplot(dtflights[, .N, by = .(month, hour)], aes(hour, month, fill = N)) + geom_tile()
```
#The average temperature at noon in JFK for each month based on the weather dataset
```{r}

dt <- data.table(weather)
ggplot(dt[origin == 'JFK', .(temp = mean(temp, na.rm = TRUE)), by = month], aes(month, temp)) +
  geom_bar(stat = 'identity', color = 'red')
```

##The average seats and engines per manufacturer
```{r}
dta <- dtplanes[, .(seats = mean(seats, na.rm = TRUE),
              engines = mean(engines, na.rm = TRUE)), by = manufacturer]
ggplot(dta, aes(seats, engines, label = manufacturer)) + geom_bar(stat = 'identity', color = 'red') + geom_text()

```

##The average windspeed and wind direction  per origin
```{r}
dta <- dtweather[, .(wind_dir = mean(wind_dir, na.rm = TRUE),
              wind_speed = mean(wind_speed, na.rm = TRUE)), by = origin]
ggplot(dta, aes(wind_speed, wind_dir, label = origin)) + geom_bar(stat = 'identity', color = 'red') + geom_text()
```

##The average temperature and humid  per origin
```{r}
dta <- dtweather[, .(temp = mean(wind_dir, na.rm = TRUE),
              humid = mean(wind_speed, na.rm = TRUE)), by = origin]
ggplot(dta, aes(temp, humid, label = origin)) + geom_bar(stat = 'identity', color = 'red') + geom_text()
```

## How to affect humidity or temperature on delays

## Featuring
```{r}
dtflights$hour <- ifelse(dtflights$hour == 24, 0, flights$hour)

flights_weather <- left_join(dtflights, dtweather)

flights_weather$arr_delay <- ifelse(flights_weather$arr_delay >= 0,
                                    flights_weather$arr_delay, 0)
flights_weather$dep_delay <- ifelse(flights_weather$dep_delay >= 0,
                                    flights_weather$dep_delay, 0)
flights_weather$total_delay <- flights_weather$arr_delay + flights_weather$dep_delay

data <- select(flights_weather, total_delay, temp, dewp, humid,
                   wind_dir, wind_speed, wind_gust, precip, pressure, visib)
```

## Effect of the humidity on delays

```{r}
g <- ggplot(data, aes(y = humid, x = total_delay, 
                          title = "Total Delay / Humidity"))
g + geom_smooth() + ylab("Relative Humidity") + 
  xlab("Total Delay (mins)")
```

## Effect of the temperature on delays

```{r}
g <- ggplot(data, aes(y = temp, x = total_delay, 
                          title = "Total Delay / Temperature"))
g + geom_smooth() + ylab("Temperature") + 
  xlab("Total Delay (mins)")
```



##Effect of the wind speed on delays

```{r}
g <- ggplot(data, aes(y = wind_speed, x = total_delay, 
                          title = "Total Delay / wind speed"))
g + geom_smooth() + ylab("wind speed") + 
  xlab("Total Delay (mins)")
```

###Average departure delay by manufacutrer

```{r}
flights_planes <- left_join(dtflights, dtplanes, by = 'tailnum')
dta <- ddply(flights_planes,~manufacturer,summarise, delay=mean(dep_delay, na.rm=TRUE))
setorder(dta, delay)

ggplot(dta, aes(manufacturer, delay)) + geom_bar(stat = 'identity') + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

###Average arrival delay by manufacutrer
```{r}
dta <- ddply(flights_planes,~manufacturer,summarise, delay=mean(arr_delay, na.rm=TRUE))
setorder(dta, delay)

ggplot(dta, aes(manufacturer, delay)) + geom_bar(stat = 'identity') + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```






###More feature engineering
Data table without NA's, make binary variable which shows if an airplane delay more than 15 minutes.
I decreased the observations number to 15.0000.

```{r}



dtflights <- subset (flights, !is.na(flights$dep_time) &
    !is.na(flights$dep_delay) &
    !is.na(flights$arr_delay) &
    !is.na(flights$dep_time) &
    !is.na(flights$arr_time))
dtflights$tailnumfac <- as.factor(dtflights$tailnum)
dtflights$carrierfac <- as.factor(dtflights$carrier)
dtflights$originfac <- as.factor(dtflights$origin)
dtflights$destfac <- as.factor(dtflights$dest)

dtflights <- dtflights[sample(1:nrow(dtflights), 15000, replace=FALSE),]

dtflights$tailnumnum <- as.numeric(dtflights$tailnumfac)
dtflights$carriernum <- as.numeric(dtflights$carrierfac)
dtflights$originnum <- as.numeric(dtflights$originfac)
dtflights$destnum <- as.numeric(dtflights$destfac)

dtflights$year <- NULL
dtflights$tail_num <- NULL
dtflights$tailnum <- NULL
dtflights$carrier <- NULL
dtflights$dest <- NULL
dtflights$origin <- NULL
dtflights$tailnum <- NULL
dtflights$tailnumfac <- NULL
dtflights$carrierfac <- NULL
dtflights$destfac <- NULL
dtflights$originfac <- NULL
dtflights$tailnumfac <- NULL

str(dtflights)

dtflights$arrdelay15 <- ifelse(dtflights$arr_delay > 15,1,0)
dtflights$year <- NULL
time_format <- "%Y-%m-%d %H:%M:%S"
dtflights$weekday <- as.factor(format(strptime(dtflights$time_hour, format=time_format),"%A"))
dtflights$time_hour <- NULL


```

## 2-Nearest Neighbors algorithm

```{r}
dtflights$rnd <-runif(dim(dtflights[1])) 
dtflights <- dtflights[order(dtflights$rnd),]


dtflights

train <- dtflights[0:round((dim(dtflights)[1])*0.7),]
test <- dtflights[(round((dim(dtflights)[1])*0.7)+1):(dim(dtflights)[1]),]
dtflights$rnd <-NULL

fit <- knn(train[,1:15], test[,1:15], train$arrdelay15, k = 2)

pander(table(test$arrdelay15,fit))
```

## 5-Nearest Neighbors algorithm

```{r}
fit2 <- knn(train[,1:15], test[,1:15], train$arrdelay15, k = 5)
pander(table(test$arrdelay15,fit2))

total <- dim(test)[1]

```
All in all the 2-NN model has provided 78 % result, the 5- NN model has provided 80 % good result.

#Modeling
### CLEAR MEMORY

```{r}
rm(list = ls())
```

```{r}
library(h2o)
h2o.init()
```


## write demo data to disk

```{r}
library(nycflights13)
write.csv(flights, 'flights.csv', row.names = FALSE)
flights.hex <- h2o.uploadFile('flights.csv', destination_frame = 'flights')
str(flights.hex)
head(flights.hex)
head(flights.hex, 3)
summary(flights.hex)
```

## convert numeric to factor/enum
```{r}

flights.hex[, 'flight'] <- as.factor(flights.hex[, 'flight'])
summary(flights.hex)
flights.hex$flight <- as.factor(flights.hex$flight)
for (v in c('month', 'day', 'dep_delay', 'arr_delay')) {
    flights.hex[, v] <- as.factor(flights.hex[, v])
}
summary(flights.hex)
```

## drop columns
```{r}
dt <- data.table(flights)
dt$delay15 <- ifelse(dt$arr_delay > 15,1,0)
str(dt)
dt <- dt[, .(month, day, dest, origin,
             carrier, flight, tailnum, distance, delay15)]
```


## transform to factor
```{r}
for (v in c('month', 'day', 'flight', 'carrier')) {
    set(dt, j = v, value = as.factor(dt[, get(v)]))
}
str(dt)
```

## re-upload to H2O
```{r}
h2o.ls()
h2o.rm('flights')
as.h2o(dt, 'flights')
```

## split the data
```{r}
flights.hex <- h2o.getFrame('flights')
h2o.splitFrame(data = flights.hex , ratios = 0.75, destination_frames = c('train', 'test'))
h2o.ls()

```

## build the first model

```{r}
flights.rf <- h2o.randomForest(
    x = names(flights.hex),
    y = 'delay15',
    training_frame = 'train',
    validation_frame = 'test')
flights.rf

```

## GBM
```{r}
flights.gbm <- h2o.gbm(
    x = names(flights.hex),
    y = 'delay15',
    training_frame = 'train',
    validation_frame = 'test',
    model_id = 'flights_gbm')
flights.gbm
```



